Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-07-23 agujournaltemplate.tex
Reworked Abstract + Plain language summary

\begin{abstract}
Phytoplankton communities play a crucial role in ocean biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems. Long-term time series of phytoplankton can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities. This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Water samples were taken during monthly cruises to determine plankton community composition and local water column conditions, which we integrated with climate indices and meteorological data. 
Using clustering algorithms, we investigate the structure of the phytoplankton community data, revealing two clusters that described a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014. However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (sea surface temperature and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.

@@

% NO MORE THAN 200 Words
% Topic Overview
Phytoplankton are microscopic algae that form the foundation of marine food webs and ecosystems. Their growth and community composition are impacted by the local physical and chemical environment, which is determined by currents and climate, for example, through processes such as wind-driven upwelling of nutrients from the deeper ocean. Time series data of phytoplankton biomass and community composition can be very helpful in untangling the link between global and local climate and its effects on the phytoplankton community.
% Paper Overview
In this study, we set out to analyze changes in the phytoplankton community between 1995 and 2017 at a time series location in a large coastal basin in Venezuela, the Cariaco Basin. Previous analysis of the data found strong shifts within the phytoplankton community that could be related to impacts of global warming. Our main question in approaching the data was whether we can confirm this shift in community composition or whether other climate processes can explain the observed patterns. 
% Findings Summary
We found that the phytoplankton community could be separated into two groups. The first group includes an early and late period (1995-2003 and 2014-2017) that was characterized by robust phytoplankton growth driven by relatively strong upwelling of nutrients. The second group is the intermediate period, where a reduction in upwelling and phytoplankton biomass was observed. Additionally, we related physical and climatic variables, such as sea surface temperature and climate indices that relate to long-term oscillations in the North Atlantic, to the changes in community composition. By allowing for a time lag, we found that the Atlantic Meridional Oscillation (AMO) index was the strongest predictor of community changes, with temperature and 
% Key Takeaways

@@

\subsection{Community reorganization and biomass shifts}
% P1: Introductory paragraph highlighting the overall dynamics/shifts
This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
These changes are well visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustering}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014. 
By considering the phytoplankton community time series until the end of 2017, the temporally complete dataset, we find that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}). Concurrently, we observe an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).

@@

The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea. During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range. Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in regime 1 (see Figure \ref{fig:clustering} $e$).

>>>>

\begin{abstract}

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series of phytoplankton can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. We considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states associated with a shift from high to low upwelling conditions in 2004, followed by a return to a community and conditions similar to those before the shift in 2014. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2 index. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilient nature of the tropical marine ecosystem under global change.

@@

% Topic Overview
Phytoplankton are microscopic algae that form the foundation of the ocean food web. Because they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local wind and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014. Interestingly, species diversity declined between 1998 and 2009, although upwelling had not yet changed. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

\subsection{Community reorganization and biomass shifts}
% P1: Introductory paragraph highlighting the overall dynamics/shifts
This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
These changes are clearly visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustcomp}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014. 
By considering the phytoplankton community time series until the end of 2017, the temporally complete dataset, we find that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}). Concurrently, we observe an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).

@@

The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea. During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range. Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in regime 1 (see Figure \ref{fig:clustering} $e$). 
- Strong El Niño events increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies \cite{huang_enso_2023}.
